---
title: "Applying Machine learning to build a predictive model of how well exercises were made"
author: "Hesham Elhalawani"
date: "June 25, 2017"
output: html_document
---

##Statement of problem

One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively.In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants.

 

##Data source

The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment. 


##Data processing

Read the training data into a data table.

```{r}
require(data.table)
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
D <- fread(url)
```

Read the testing data into a data table.

```{r}
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
DTest <- fread(url)
```


Excluding variables with missing values

Belt, arm, dumbbell, and forearm variables that do not have any missing values in the test dataset will be **predictor candidates**.

```{r}
isAnyMissing <- sapply(DTest, function (x) any(is.na(x) | x == ""))
isPredictor <- !isAnyMissing & grepl("belt|[^(fore)]arm|dumbbell|forearm", names(isAnyMissing))
predCandidates <- names(isAnyMissing)[isPredictor]
predCandidates
```

Subset the primary dataset to include only the **predictor candidates** and the outcome variable, `classe`.

```{r}
varToInclude <- c("classe", predCandidates)
D <- D[, varToInclude, with=FALSE]
dim(D)
names(D)
```

Make `classe` into a factor.

```{r}
D <- D[, classe := factor(D[, classe])]
D[, .N, classe]
```

Split the dataset into a 60% training and 40% probing dataset.

```{r}
library(caret)
seed <- as.numeric(as.Date("2014-10-26"))
set.seed(seed)
inTrain <- createDataPartition(D$classe, p=0.6)
DTrain <- D[inTrain[[1]]]
DProbe <- D[-inTrain[[1]]]
```

Preprocess the prediction variables by centering and scaling.

```{r}
X <- DTrain[, predCandidates, with=FALSE]
preProc <- preProcess(X)
preProc
XCS <- predict(preProc, X)
DTrainCS <- data.table(data.frame(classe = DTrain[, classe], XCS))
```

Apply the centering and scaling to the probing dataset.

```{r}
X <- DProbe[, predCandidates, with=FALSE]
XCS <- predict(preProc, X)
DProbeCS <- data.table(data.frame(classe = DProbe[, classe], XCS))
```

Check for near zero variance.

```{r}
nzv <- nearZeroVar(DTrainCS, saveMetrics=TRUE)
if (any(nzv$nzv)) nzv else message("No variables with near zero variance")
```

Examine groups of prediction variables.

```{r histGroup}
histGroup <- function (data, regex) {
  col <- grep(regex, names(data))
  col <- c(col, which(names(data) == "classe"))
  require(reshape2)
  n <- nrow(data)
  DMelted <- melt(data[, col, with=FALSE][, rownum := seq(1, n)], id.vars=c("rownum", "classe"))
  require(ggplot2)
  ggplot(DMelted, aes(x=classe, y=value)) +
    geom_violin(aes(color=classe, fill=classe), alpha=1/2) +
#     geom_jitter(aes(color=classe, fill=classe), alpha=1/10) +
#     geom_smooth(aes(group=1), method="gam", color="black", alpha=1/2, size=2) +
    facet_wrap(~ variable, scale="free_y") +
    scale_color_brewer(palette="Spectral") +
    scale_fill_brewer(palette="Spectral") +
    labs(x="", y="") +
    theme(legend.position="none")
}
histGroup(DTrainCS, "belt")
histGroup(DTrainCS, "[^(fore)]arm")
histGroup(DTrainCS, "dumbbell")
histGroup(DTrainCS, "forearm")
```


# Train a prediction model

Using random forest, the out of sample error should be small.
The error will be estimated using the 40% probing sample.
I would be quite happy with an error estimate of 3% or less.

Set up the parallel clusters.

```{r}
library(parallel)
library(doParallel)
cl <- makeCluster(detectCores() - 1)
registerDoParallel(cl)
```

Set the control parameters.

```{r}
ctrl <- trainControl(classProbs=TRUE,
                     savePredictions=TRUE,
                     allowParallel=TRUE)
```

Fit model over the tuning parameters.

```{r}
method <- "rf"
system.time(trainingModel <- train(classe ~ ., data=DTrainCS, method=method))
```

Stop the clusters.

```{r}
stopCluster(cl)
```

## Evaluate the model on the training dataset

```{r}
trainingModel
hat <- predict(trainingModel, DTrainCS)
confusionMatrix(hat, DTrain[, classe])
```

## Evaluate the model on the probing dataset

```{r}
hat <- predict(trainingModel, DProbeCS)
confusionMatrix(hat, DProbeCS[, classe])
```

## Display the final model

```{r finalModel}
varImp(trainingModel)
trainingModel$finalModel
```

**The estimated error rate is less than 1%.**

Save training model object for later.

```{r}
save(trainingModel, file="trainingModel.RData")
```


# Predict on the test data

Load the training model.

```{r}
load(file="trainingModel.RData", verbose=TRUE)
```

Get predictions and evaluate.

```{r}
DTestCS <- predict(preProc, DTest[, predCandidates, with=FALSE])
hat <- predict(trainingModel, DTestCS)
DTest <- cbind(hat , DTest)
subset(DTest, select=names(DTest)[grep("belt|[^(fore)]arm|dumbbell|forearm", names(DTest), invert=TRUE)])
```

## Submission to Coursera

Write submission files to `predictionAssignment_files/answers`.

```{r}
pml_write_files = function(x){
  n = length(x)
  path <- "predictionAssignment_files/answers"
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=file.path(path, filename),quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(hat)
```